Aa Language Reference Manual

### 2016-10-23 v2/Ahir-asm/devel/docs/AaLRM.tex
Added concept of parameters to allow abstraction of magic numbers to meaningful symbols/expresions.

Variables in  the program can belong to the following classes:
\begin{itemize}
\item {\bf Storage} variables declared in the program
can be written into from multiple
points in the program.  When declared at the program
scope, storage variables are visible throughout
the program.  Storage variables can be thought of 
as memory locations which can be accessed from different
points in the program.
A storage variable declaration has the form
\begin{verbatim}
$storage <var-name>: <var-type> 
\end{verbatim}
For example
\begin{verbatim}
$storage mempool: $array [1024] $of $uint<32>
\end{verbatim}
\item {\bf Pipe} variables declared at the program scope
are globally visible buffers (which
can be defined to be either first-in-first-out (FIFO) or
last-in-first-out (LIFO)) that can 
be read from anywhere in the program and can be
written into from anywhere in th program.  A pipe
is a FIFO (or FIFO) buffer with a specified number of
slots.  A write to a pipe succeeds
only if there is an empty slot in the FIFO (a successful
write uses up an empty slots.  A read
from the pipe succeeds only if there is at
least one occupied slot in the pipe (a successful read
creates an empty slot in the pipe). 
Thus, pipes can be used to model queues as
well as stacks, and also provide a mechanism for synchronization and
for message passing in an \Aa program.
\begin{itemize}
\item A pipe that is only written into in the program, but
is not read from, is assumed to have a destination outside
the program.  Such a pipe is an output port of the
program.
\item A pipe that is only read from in a program but not
written into, is assumed to have a source outside the program.
Such a pipe is an input port of the program.
\end{itemize}
A pipe declaration has the form
\begin{verbatim}
($lifo | $noblock)? $pipe <pipe-name>: <pipe-type>  ($depth <integer>)?
\end{verbatim}
For example
\begin{verbatim}
$pipe myfifo: $uint<32> $depth 16 // 16-slot queue
$lifo $pipe mystack: $uint<32> $depth 4 // 4-slot stack
$noblock $pipe cmd_pipe: $uint<8> // non-blocking reads.
\end{verbatim}
A non-blocking pipe when read, will deliver $0$ if
it does not contain data (thus, you will have to reserve
the $0$ for invalid data).

Pipe variable declarations can also be appended by
the following keywords:
\begin{verbatim}
   ($in | $out)? ($signal)? 
\end{verbatim}
These additional keywords mean the following things:
\begin{itemize}
\item Inclusion of ``\$in'' (respectively ``\$out'') means
that the pipe is a read-only (respectively write-only) object.
\item The ``\$signal'' keyword means that the pipe behaves
like a variable.  That is, it remembers the last value
that was written to it.    
\end{itemize}
\item {\bf Constant} variables declared at
the program scope are globally visible and have
initial values which can never be altered in the program.
Thus, an assignment to a constant is an error.
A constant declaration has the form
\begin{verbatim}
$constant <constant-name>: <constant-type> := <initial-value>
\end{verbatim}
For example
\begin{verbatim}
$constant wordlength: $uint<32> := 16
\end{verbatim}
\item {\bf Implicit} or {\bf Static Single Assignment}
variables are created by statements in the \Aa program
(as described below in Section \ref{sec:Statements}).
These are analogous to registers that can be written
into only from one point,  but can be
read from several points in the program.
Implicit variables are not declared, but are defined
by the statements which write to them.
\end{itemize}

@@

@@

Scalar types can be one of 
\begin{itemize}
\item Unsigned integers:  An unsigned integer type
has a width parameter and is specified as 
\begin{verbatim}
$uint<width>
\end{verbatim}
The width parameter can be any positive number.
Values corresponding to this type are to be
viewed as unsigned integers represented by 
a binary sequence of the specified width.
\item Signed integers:  A signed integer type
has a width parameter and is specified as 
\begin{verbatim}
$int<width>
\end{verbatim}
The width parameter can be any positive number.
Values corresponding to this type are to be
viewed as integers maintained in the
two-s complement form  by 
a binary sequence of the specified width.

\item Pointers:  A pointer is an unsigned
integer with a default pointer width
(set to 32 for now), which specifies the
type of object to which it points.
For example
\begin{verbatim}
$pointer< $uint<32> > 
\end{verbatim}
is a pointer which refers to a storage
object of type \begin{verbatim}$uint<32>\end{verbatim}
\item Floats: A float is parametrized by
two integers, the width of the exponent,
and the width of the mantissa.  
The specification is
\begin{verbatim}
$float<exponent,mantissa>
\end{verbatim} %$
where the exponent and mantissa must be positive
integers.  The float is represented by 
a word with $exponent+mantissa+1$ bits
(with the additional bit needed for the sign).
The standard IEEE 754 float and double
precision representations correspond
to
\begin{verbatim}
$float<8,23> 
\end{verbatim}
and
\begin{verbatim}
$float<11,52> 
\end{verbatim}
respectively.  Currently, these are the
only float types that are supported.
\end{itemize}

@@

Slice expressions can be used to extract a contiguous
bit-field from a value.  They have the
following syntax:
\begin{verbatim}
LPAREN $slice <expression> <high-index> <low-index> RPAREN
\end{verbatim}
where high-index and low-index are constant non-negative
integers.  The result of executing the expression is
an unsigned integer whose width is $(high-index - low-index)+1$.
The value produced by expression may be of any type, but
for the slicing, it is flattened in the natural manner
(see the notes about aggregate types).  All bit-vectors
are assumed to indexed as $HIGH$ dowto $LOW$.

@@

For example
\begin{verbatim}
( $slice ($bitcast ($uint<4>) _h3) 0 3 1 2 2 1 3 0 ) = _he
\end{verbatim}

>>>>

Variables in  the program can belong to the following classes:
\begin{itemize}
\item {\bf Parameters} can be used to configure the Aa program to
a certain extent.  These are 32-bit integers declared as
\begin{verbatim}
$parameter PIPEDEPTH 16
$parameter PIPEDEPTH2  (2 * PIPEDEPTH) 
\end{verbatim}
for example.  
Parameters can be used to specify some 
constants inside
the program (this will be indicated whenever relevant).
Parameters
should be known at the point at which they are used.  Also,
parameters can be declared only at global program scope (that is
not inside a module).
\item {\bf Storage} variables declared in the program
can be written into from multiple
points in the program.  When declared at the program
scope, storage variables are visible throughout
the program.  Storage variables can be thought of 
as memory locations which can be accessed from different
points in the program.
A storage variable declaration has the form
\begin{verbatim}
$storage <var-name>: <var-type> 
\end{verbatim}
For example
\begin{verbatim}
$storage mempool: $array [1024] $of $uint<32>
\end{verbatim}
\item {\bf Pipe} variables declared at the program scope
are globally visible buffers (which
can be defined to be either first-in-first-out (FIFO) or
last-in-first-out (LIFO)) that can 
be read from anywhere in the program and can be
written into from anywhere in th program.  A pipe
is a FIFO (or FIFO) buffer with a specified number of
slots.  A write to a pipe succeeds
only if there is an empty slot in the FIFO (a successful
write uses up an empty slots.  A read
from the pipe succeeds only if there is at
least one occupied slot in the pipe (a successful read
creates an empty slot in the pipe). 
Thus, pipes can be used to model queues as
well as stacks, and also provide a mechanism for synchronization and
for message passing in an \Aa program.
\begin{itemize}
\item A pipe that is only written into in the program, but
is not read from, is assumed to have a destination outside
the program.  Such a pipe is an output port of the
program.
\item A pipe that is only read from in a program but not
written into, is assumed to have a source outside the program.
Such a pipe is an input port of the program.
\end{itemize}
A pipe declaration has the form
\begin{verbatim}
($lifo | $noblock)? $pipe <pipe-name>: <pipe-type>  ($depth <integer>)?
\end{verbatim}
For example
\begin{verbatim}
$pipe myfifo: $uint<32> $depth 16 // 16-slot queue
$lifo $pipe mystack: $uint<32> $depth 4 // 4-slot stack
$noblock $pipe cmd_pipe: $uint<8> // non-blocking reads.
\end{verbatim}
A non-blocking pipe when read, will deliver $0$ if
it does not contain data (thus, you will have to reserve
the $0$ for invalid data).

Pipe variable declarations can also be appended by
the following keywords:
\begin{verbatim}
   ($in | $out)? ($signal)? 
\end{verbatim}
These additional keywords mean the following things:
\begin{itemize}
\item Inclusion of ``\$in'' (respectively ``\$out'') means
that the pipe is a read-only (respectively write-only) object.
\item The ``\$signal'' keyword means that the pipe behaves
like a variable.  That is, it remembers the last value
that was written to it.    
\end{itemize}
Pipe depths can be specified by parameters or parameter
expressions.
\item {\bf Constant} variables declared at
the program scope are globally visible and have
initial values which can never be altered in the program.
Thus, an assignment to a constant is an error.
A constant declaration has the form
\begin{verbatim}
$constant <constant-name>: <constant-type> := <initial-value>
\end{verbatim}
For example
\begin{verbatim}
$constant wordlength: $uint<32> := 16
\end{verbatim}
\item {\bf Implicit} or {\bf Static Single Assignment}
variables are created by statements in the \Aa program
(as described below in Section \ref{sec:Statements}).
These are analogous to registers that can be written
into only from one point,  but can be
read from several points in the program.
Implicit variables are not declared, but are defined
by the statements which write to them.
\end{itemize}

@@

Note that the pipeline depth and buffering can be
specified as parameters or parameter expressions.

@@

Scalar types can be one of 
\begin{itemize}
\item Unsigned integers:  An unsigned integer type
has a width parameter and is specified as 
\begin{verbatim}
$uint<width>
\end{verbatim}
The width can be any positive integer (or a parameter
expression that evaluated to a positive integer).
Values corresponding to this type are to be
viewed as unsigned integers represented by 
a binary sequence of the specified width.
\item Signed integers:  A signed integer type
has a width parameter and is specified as 
\begin{verbatim}
$int<width>
\end{verbatim}
The width parameter can be any positive number.
Values corresponding to this type are to be
viewed as integers maintained in the
two-s complement form  by 
a binary sequence of the specified width.

\item Pointers:  A pointer is an unsigned
integer with a default pointer width
(set to 32 for now), which specifies the
type of object to which it points.
For example
\begin{verbatim}
$pointer< $uint<32> > 
\end{verbatim}
is a pointer which refers to a storage
object of type \begin{verbatim}$uint<32>\end{verbatim}
\item Floats: A float is parametrized by
two integers, the width of the exponent,
and the width of the mantissa.  
The specification is
\begin{verbatim}
$float<exponent,mantissa>
\end{verbatim} %$
where the exponent and mantissa must be positive
integers (or parameter expressions that evaluate
to positive integers).  The float is represented by 
a word with $exponent+mantissa+1$ bits
(with the additional bit needed for the sign).
The standard IEEE 754 float and double
precision representations correspond
to
\begin{verbatim}
$float<8,23> 
\end{verbatim}
and
\begin{verbatim}
$float<11,52> 
\end{verbatim}
respectively.  Currently, these are the
only float types that are supported.
\end{itemize}

@@

Slice expressions can be used to extract a contiguous
bit-field from a value.  They have the
following syntax:
\begin{verbatim}
LPAREN $slice <expression> <high-index> <low-index> RPAREN
\end{verbatim}
where high-index and low-index are constant non-negative
integers (or parameter expressions that evaluate to
non-negative integers).  The result of executing the expression is
an unsigned integer whose width is $(high-index - low-index)+1$.
The value produced by expression may be of any type, but
for the slicing, it is flattened in the natural manner
(see the notes about aggregate types).  All bit-vectors
are assumed to indexed as $HIGH$ dowto $LOW$.

@@

For example
\begin{verbatim}
( $bitmap ($bitcast ($uint<4>) _h3) 0 3 1 2 2 1 3 0 ) = _he
\end{verbatim}

Note: this expression is not currently supported in hardware,
but will eventually be, maybe.
